Best Practices in Latent Class Analysis using Free Open Source Software

### 2023-02-16 manuscript.tex
Update pt. 1
Updated manuscript and cleaned references

\shorttitle{BEST PRACTICES LATENT CLASS ANALYSIS}

@@

\abstract{%
Latent class analysis (LCA) refers to a family of techniques for identifying groups
in data based on a parametric model.
Examples include mixture models, LCA with ordinal
indicators, and latent class growth analysis.
Despite the popularity of this technique, there is limited
guidance with respect to best practices in estimating and reporting mixture
models. Moreover, although user-friendly interfaces for advanced mixture
modeling have long been available in commercial software packages, open
source alternatives have remained somewhat inaccessible. This tutorial
describes best practices for the estimation and reporting of LCA,
using free and open source software in R. To this end, this tutorial
introduces new functionality for estimating and reporting mixture
models in the \texttt{tidySEM} R-package whose backend relies on \texttt{OpenMx}.
}

@@

Latent class analysis (LCA) is an umbrella term that refers to a number
of techniques for estimating unobserved group membership based on a
parametric model of one or more observed indicators of group membership.
The types of LCA have become quite popular across scientific fields,
most notably finite Gaussian mixture modeling and latent profile
analysis. Jeroen K. Vermunt et al. (2004) defined LCA more generally as virtually
any statistical model where ``some of the parameters {[}\ldots{]} differ across
unobserved subgroups''.

Despite the popularity of LCA, two challenges impede broader adoption
and correct application of the method. The first challenge is that most
existing software that implements these models is commercial and
closed-source. The second challenge is that there is a lack of standards
for estimating and reporting LCA, and some myths and misunderstandings
have been perpetuated in prior literature. This introduces a risk of
misapplications of the technique, and complicates manuscript review and
assessment of the quality of published studies. The present paper seeks
to address these challenges by introducing updated guidelines for
estimating and reporting LCA, based on current best practice.
Furthermore, meeting the increasing demand for open source research
software, this paper introduces new functionality in the R-package
\texttt{tidySEM} (\textbf{vanlissaTidySEMTidyStructural2022?}) to perform LCA using free
open source software (FOSS).

@@

Despite the popularity of LCA, most existing software is commercial and
closed-source, for example, Mplus (\textbf{muthenMplusUserGuide1998?}), for which
the R-package \texttt{MplusAutomation} offers augmented LCA functionality
(\textbf{hallquistMplusAutomationPackageFacilitating2018?}), and LatentGOLD
{[}REF{]}.
A
downside of commercial software is that its use is restricted to
individuals and institutions who can afford a license. Another downside
is that the proprietary source code cannot be audited, debugged, or
enhanced by third parties. This goes against open science principles
(\textbf{lamprechtFAIRPrinciplesResearch2019?}), incurs a risk that mistakes in
the source code go unnoticed, and curbs progress as independent
developers cannot fix bugs or add functionality.

There are several compelling reasons to prefer free open source software
(FOSS) instead. FOSS is freely available for anyone to use, modify, and
distribute, which means that scientists can access and use a wide range
of powerful tools without financial impediments. This is invaluable for
reducing gatekeeping of researchers with limited budgets, such as those
from developing countries. Furthermore, FOSS promotes collaboration
within the scientific community, as its users can contribute to the
development, maintenance, and support of the software. Finally, FOSS
promotes transparency and reproducibility: Because the source code for
FOSS is accessible, researchers can review and verify the methods and
algorithms used, which helps to ensure the integrity and reliability of
research findings. Overall, FOSS has the potential to significantly
improve the efficiency and effectiveness of scientific research, while
also promoting transparency, collaboration, and inclusion within the
scientific community.

Notable FOSS for LCA includes mclust (Scrucca, Fop, Murphy, \& Raftery, 2016), depmixS4
(Visser \& Speekenbrink, 2010), and OpenMx (Neale et al., 2016). A crucial limitation of
these existing packages is that some have limited functionality (e.g.,
Mclust), or implement specific innovations (e.g., depmixS4). OpenMx is
unique in that, like Mplus, it implement LCA in the context of a fully
featured structural equation modeling framework. Remaining obstacles to
the use of \texttt{OpenMx} is that it is not very user-friendly and its
functionality for LCA is poorly documented.

These limitations are addressed by the \texttt{tidySEM} package, which provide
a user-friendly interface for LCA in \texttt{OpenMx}. The package also includes
convenience functions to obtain relevant statistics, tests and plots in
accordance with best practices in estimation and reporting LCA, as
described in this paper. The primary purpose of the \texttt{tidySEM} package is
to provide a tidy workflow for generating, estimating, reporting, and
plotting structural equation models in a software-agnostic way. The user
interface is simple, as models can be specified using the widely used
\texttt{lavaan} syntax (Rosseel, 2012). The package contains
user-friendly wrapper functions to conduct LCA with sensible defaults,
but the resulting models can still be fully customized (low barrier,
high ceiling). Specifically, LCA models constructed in \texttt{tidySEM} inherit
from \texttt{OpenMx}' format \texttt{MxModel}, and remain compatible with downstream
functions that process such models. Being FOSS, \texttt{tidySEM} facilitates
fully-reproducible analyses and contributes to open science. It was
developed with open science in mind, and adheres to the FAIR software
principles (Findable, Accessible, Interoperable, and Reusable, \textbf{lamprechtFAIRPrinciplesResearch2019?}), and the \href{https://bestpractices.coreinfrastructure.org/en/criteria}{OpenSSF Best
Practices} for
FOSS software.

@@

LCA refers to methods for estimating unobserved groups based on a
parametric model of observed indicators of group membership
(\textbf{vermuntLatentClassAnalysis2004?}). It has become popular across
scientific fields, under a number of different names; most notably
(finite Gaussian) mixture modeling and latent profile analysis. The
concept of LCA can be understood in different ways. Generally speaking,
LCA assumes that the study population is composed of \(K\) subpopulations.
It further assumes that the observed data are a mixture of data
originating from those subpopulations; hence the name mixture model.
Consider the simplest possible univariate ``model'', which is a normal
distribution. This model has two parameters: the mean and the variance.
LCA aims to estimate the values of those parameters across a known
number of classes \(K\), as well as the probability of every individual
\(i\) belonging to classes \(1 \ldots k\). Commonly, the same model is
estimated across all classes, but with different parameters for each
class (i.e., class-specific means and variances).

@@

The distribution is evidently bimodal, which bodes well for the intended
mixture model. In this case, the number of classes is known a-priori.
When estimating a two-class mixture model, the detective observes that
the model estimates the mean shoe size of the two groups are equal to
9.22 and
7.25, which is close to the true
means of the two groups, namely 9.04
and 6.93. When tabulating estimated
group membership against observed (known) group membership, it can be
seen that women are classified with a high degree of accuracy, but men
are not:

@@

Another way to conceptualize LCA is by analogy to a measurement model in
structural equation modeling (\textbf{molenaarArbitraryNatureLatent1994?}). A
mixture model is like confirmatory factor analysis, except with a
categorical instead of a continuous latent variable. One difference
between the two techniques is that factor analysis can be considered as
a way to group observed \emph{variables} into latent constructs, and LCA
groups \emph{individuals} into classes. While factor analysis seeks to
describe the relations between variables, LCA seeks to describe
unobserved groups and classify individual observations to those groups.
In
line with this distinction, LCA (mixture models) is referred to as a
``person-centered'' technique, and factor analysis as a
``variable-centered'' technique (Masyn, 2013; Nylund-Gibson \& Choi, 2018).

When the focus is on the model parameters in each group, LCA can be
thought of as similar to a multi-group structural equation model. The
main distinction is that group membership is not known a-priori, but
estimated from the data. Whereas in a multi-group model, the data are
split by group and treated as independent samples, in LCA, all cases
contribute to the estimation of all parameters in all groups. The
relative contribution of each case to the parameters of each group is
determined by that case's posterior probability of belonging to that
group.

When the focus is on each individual's estimated class membership, LCA
can be thought of as a type of clustering algorithm that corrects for
(random) measurement error. Specifically, the posterior class
probability that an individual belongs to a latent class can be computed
from the likelihood of that individual's observed data under given the
class-specific model. These posterior class probabilities can be used to
weight follow-up analyses, or individuals can be assigned to classes
based on their highest class probability. If the classes are clearly
distinct, measurement error is low, and the latter approach may be
acceptable. This perspective on mixture modeling is sometimes referred
to as ``model-based clustering'' (Hennig, Meila, Murtagh, \& Rocci, 2015; Scrucca et al., 2016). Many clustering algorithms apply some recursive
splitting algorithm to the data. By contrast ``model-based'' clustering
refers to the fact that LCA estimates cluster membership based on a
parametric model. Since a parametric model estimates only a fixed number
of parameters, it can be a relatively parsimonious solution compared to
non-parametric techniques.

@@

The term LCA was initially conceived to refer to analyses with
categorical, usually binary indicators (Collins \& Lanza, 2009). It was
later generalized to refer to any technique that estimates a categorical
latent variable based on a parametric model of observed indicators, as
it only refers to the purpose of the analysis, and does not imply
restrictions to the model used, or the level of measurement of the
indicators (Jeroen K. Vermunt et al., 2004). To prevent ambiguity, we refer to LCA
with binary or ordinal indicators as \emph{LCA with ordinal indicators}.
Given the abundance of redundant terms in use for closely related types
of LCA models, we here provide a rudimentary taxonomy.

A common type of LCA is the \emph{finite Gaussian mixture model}; a
univariate analysis where the observed distribution of a single
continuous variable is assumed to result from a mixture of a known
number of normal (Gaussian) distributions. The parameters of a finite
Gaussian mixture model are the means and variances of these underlying
normal distributions. The analysis of shoe sizes presented earlier is a
canonical example of this type of analysis. In the multivariate case,
with more than one indicator variable, the parameters of a mixture model
are the means, variances, and covariances between the indicators (which
can be standardized to obtain correlations). These parameters can be
estimated freely, or set to be constrained across classes
(Collins \& Lanza, 2009).

The technique known as \emph{latent profile analysis (LPA)} is a special case
of the mixture model, which assumes conditional independence of the
indicators. Conditional independence means that, after class membership
is accounted for, the residual covariances/correlations between
indicators are assumed to be zero. In some cases, such constraints will
be inappropriate; for instance when the cohesion between indicators is
expected to differ between classes (Collins \& Lanza, 2009). Consider, for
example, a mixture model of ocean plastic particles, which found two
classes of particles based on length and width: A class of larger
particles with a low correlation between length and width, and a class
of smaller particles with a high correlation between length and width.
The reason for this difference in correlations makes theoretical sense:
the large particles were heterogenous in shape, whereas the smaller
particles had been polished to a more uniform (rounded) shape by the
elements (Alkema, Van Lissa, Kooi, \& Koelmans, 2022).

A longitudinal extension of LCA is the Hidden Markov Model (HMM, also
called Latent Transition Analysis), which estimates the change in class
membership over time with the markov assumption over time. The markov
assumption states that the class membership at time \(t\) is only affected
by the class membership at time \(t-1\). The structural part of the model
is comprise of the initial states (proportion of the sample in each
class at the first time point), and transition probabilities
(probabilities to switch class membership across time). The transition
probabilities are estimated as multinomial logistic regressions between
class memberships (Jeroen K. Vermunt, 2010).

Finally, some LCA models use latent variables as indicators. A common
application of this approach is in longitudinal research with repeated
measures of a construct. Examples of this approach include \emph{growth
mixture models} (GMM) and \emph{latent class growth analyses} (LCGA)
(Jung \& Wickrama, 2008). These techniques estimate a latent growth
model to describe grouped trajectories over time. The indicators of
class membership in a GMM are the intercepts, and variances, and
covariances of the latent growth variables that describe individual
trajectories on the observed variables. For linear trajectories, these
latent variables are an intercept and slope. The free variances allow
trajectories to vary within classes. The LCGA differs from the GMM in
that it assumes homogeneity of trajectories within classes. This is
achieved by restricting the (co)variances of the growth parameters to
zero. Any variance in the indicators not explained by the class-specific
latent trajectories is assumed to be error variance.

@@

Although LCA is often discussed as an exploratory analysis technique, it
can also be used in a confirmatory manner. Given that LCA is similar to
confirmatory factor analysis, it can be used to similar ends: To assess
whether a theoretical measurement model holds. This use case is relevant
when testing a theory that postulates the existence of a categorical
latent variable. For example, \emph{identity status theory} posits that, at
any given point in time, adolescents reside in one of four identity
statuses (Marcia, 1966). LCA can be used to identify these
four statuses based on observed indicators (e.g., self-reported identity
exploration and commitment)
(\textbf{luyckxDevelopmentalTypologiesIdentity2008?}). Similarly, \emph{personality
type theory} states that dimensional differences in personality can
largely be explained by an undercontrolled, overcontrolled, and
resilient type, which could also be restored using LCA
(\textbf{donnellanResilientOvercontrolledUndercontrolled2010?}).

@@

Appendix A is a tutorial to confirmatory LCA in \texttt{tidySEM}, based on a
recent study that tested the theory that identities of Belgian
adolescents with migration backgrounds could be summarized as those
relating to their heritage, national and regional identity
(Maene, D'hondt, Van Lissa, Thijs, \& Stevens, 2022).

@@

Another use case is to explore whether a population is comprised of
latent subpopulations, and if so, how many. The primary focus in this
use case is on class enumeration. For example,
one study explored heterogeneity in the population of
military service members and veterans with prior suicide attempts, and
identified two latent profiles characterized by high versus low suicide
risk based on self-reported suicidal ideation and information about
prior attempted methods and severity (Gromatsky et al., 2022).
Knowing that the population of suicidal service members is heterogeneous
can inform future research, policy, and care provided. Appendix B offers
a tutorial for exploratory LCA, based on the aforementioned study of the
distribution of ocean microplastics (Alkema et al., 2022).

@@

LCA can also be used as a measurement model, similar to confirmatory
factor analysis. This is useful when class membership is measured
imperfectly by several indicators: LCA can partial out measurement error
and restore most likely class membership. For example, one study found
discrepancies between self-reported employment status and official
register data (Pavlopoulos \& Vermunt, 2015). While it might be tempting
to assume that register data is free of error, this assumption was shown
to be false. A mixture model was used to estimate employment status,
correcting for measurement error, from both self-report and registry
data. The study investigated random and systematic measurement error in
both indicators, which reveal strengths and weaknesses of both types of
assessment.

This example also illustrates that LCA allows us to evaluate the
reliability of different indicators (Geiser \& Wurpts, 2014).
A
continuous indicator that discriminates well between classes will show
large differences in class-specific means; an ordinal indicator will
show very high or very low conditional response probabilities for some
classes but not for others. This information can be used to select the
indicators that are most diagnostic of class membership. For an example
of LCA with a focus on differential item functioning, consider a study
on university admission tests of Saudi Arabian students
(Tsaousis, Sideridis, \& AlGhamdi, 2020).

@@

LCA can also be used to reduce high-dimensional data to just a few
prototypes. Factor analysis is also sometimes used for dimension
reduction, but only accounts for linear covariance between indicators,
whereas LCA can accommodate complex - but discrete - patterns. For
example, imagine that there exist two indicators - \(x\) and \(y\), and some
individuals score high on both, some score low on both, and some score
high on \(x\) but low on \(y\). Factor analysis would only be able to
capture the differences between the first two groups well; LCA captures
differences between all three. Note that this is a pragmatic use of LCA;
there is no theory about the existence of a categorical latent variable
(cf.~Testing theory). Consequently, the class solution should be treated
as descriptive of the sample, and not be ``reified'' - treated as evidence
of the existence of a categorical latent variable.

The use of LCA for dimension reduction is common in longitudinal
research, where one developmental process is summarized using latent
class growth analysis (LCGA), and the resulting categories are then used
as a moderator of another developmental process in a multi-group model.
For example, one study conducted a LCGA of adolescent empathy
development on two dimensions of empathy and found three groups: high,
average, and low empathy (Van Lissa et al., 2015). These groups were
then used as moderator of a latent growth analysis of adolescent- and
parent-reported conflict. Results showed that the high-empathy group
evidenced greater adolescent-parent agreement about the incidence of
conflict than the other groups, and that the low empathy group had more
conflict with parents than the other groups. Note that both these key
findings are non-linear; modeling them in a single-group model would be
difficult. LCGA greatly simplifies model specification and
interpretation, albeit at a cost of some loss of information and,
potentially, generalizability. Appendix C is a tutorial on LCGA, based
on ongoing research on trajectories of depressive symptoms in military
service members after deployment.

The need for dimension reduction can be especially pertinent when using
ordinal indicators. For example, one study measured fifteen dichotomous
health indicators among injured military personnel
(MacGregor et al., 2021). Combinatorics informs us that fifteen
dichotomous items have \(2^{15}\) or \(32,768\) unique symptom combinations.
However, LCA was able to reduce response patterns to five clinically
meaningful latent classes.

@@

Another use case is to estimate individuals' unobserved class membership
based on observed indicators. The shoe size example in Figure
\ref{fig:shoedens} is a rudimentary illustration of this application.
In applied research, LCA is often used for classification in clinical
contexts. For example, LCA was used to determine clinical cut-off scores
for diagnosing whooping cough (Baughman, Bisgard, Lynn, \& Meade, 2006). Another study
used LCA to not only classify the sample used to estimate the model, but
also future cases {[}REF Zegwaard et al, in prep{]}. The study identified
four types of care providers among those who supported close kin with
mental health problems. Importantly, the LCA model was embedded in an
interactive app that could be used by healthcare professionals to
determine the most likely class membership for new care providers, to
provide more targeted support. To obtain predicted class membership
based on an existing LCA model, the authors computed the likelihood of
data from a new participant under the multivariate normal distributions
estimated within all latent classes.
These applications of LCA aid the clinical decision making process, and
help decide whether additional support is warranted, or what type of
intervention is appropriate.

@@

The matrix \(P\) can be summarized in different way. For example, the
estimated frequency table of the latent categorical variable can be
obtained by summing the posterior probabilities by class (i.e., the
column sums of \(P\)). This frequency table \(F_P\) is obtained by calling
\texttt{class\_prob(res,\ type\ =\ "sum.posterior")}. These numbers are fractional
because this table takes classification error into account. Individuals
can contribute partially to multiple classes.

>>>>

\shorttitle{BEST PRACTICES IN LATENT CLASS ANALYSIS}

@@

\abstract{%
Latent class analysis (LCA) refers to a family of techniques for identifying groups
in data based on a parametric model.
Examples include mixture models, LCA with ordinal
indicators, and latent class growth analysis.
Despite the popularity of this technique, there is limited
guidance with respect to best practices in estimating and reporting mixture
models. Moreover, although user-friendly interfaces for advanced mixture
modeling have long been available in commercial software packages, open
source alternatives have remained somewhat inaccessible. This tutorial
describes best practices for the estimation and reporting of LCA,
using free and open source software in R. To this end,
we introduce new functionality for estimating and reporting mixture models
using the \texttt{tidySEM} R-package.
}

@@

Latent class analysis (LCA) is an umbrella term that refers to a number
of techniques for estimating unobserved group membership based on a
parametric model of one or more observed indicators of group membership.
The types of LCA have become quite popular across scientific fields,
most notably finite Gaussian mixture modeling and latent profile
analysis. Vermunt et al. (2004) defined LCA more generally as
virtually any statistical model where ``some of the parameters {[}\ldots{]}
differ across unobserved subgroups''.

Despite the popularity of LCA,
two challenges impede broader adoption and correct application of the method.
The first challenge is that most existing software that implements these models is commercial and closed-source.
The second challenge is that there is a lack of standards for estimating and reporting LCA, and some myths and misunderstandings have been perpetuated in prior literature.
This introduces a risk of misapplications of the technique,
and complicates manuscript review and assessment of the quality of published studies.
The present paper seeks to address these challenges by introducing updated guidelines for estimating and reporting LCA,
based on current best practices.
Furthermore, meeting the increasing demand for open source research software,
this paper introduces new functionality in the R-package \texttt{tidySEM} (Van Lissa, 2022b) to perform LCA using free open source software (FOSS).

@@

Despite the popularity of LCA,
most existing software is commercial and closed-source,
for example, \emph{Mplus} (Muthén \& Muthén, 1998), for which the R-package \texttt{MplusAutomation} offers augmented LCA functionality (Hallquist, Wiley, \& Van Lissa, 2018), and \emph{Latent GOLD} (Vermunt \& Magidson, 2000).
A downside of commercial software is that its use is restricted to individuals and institutions that can afford a license.
Another downside is that proprietary source code cannot be audited, debugged, or enhanced by third parties.
This goes against open science principles (Lamprecht et al., 2020),
incurs a risk that mistakes in the source code may go unnoticed,
and curbs progress as independent developers cannot fix bugs or add functionality.

There are several compelling reasons to prefer free open source software instead.
FOSS is freely available for anyone to use, modify, and distribute, which means that scientists can access and use a wide range of powerful tools without financial impediments.
This is invaluable for reducing gatekeeping of researchers with limited budgets, such as those from developing countries.
Furthermore, FOSS promotes collaboration within the scientific community,
as its users can contribute to the development, maintenance, and support of the software.
Finally, FOSS promotes tranparency and reproducibility: Because the source code for FOSS is accessible,
researchers can review and verify the methods and algorithms used,
which helps to ensure the integrity and reliability of research findings.
Overall, FOSS has the potential to significantly improve the efficiency and effectiveness of scientific research, while also promoting transparency, collaboration, and inclusivity within the scientific community.

Notable FOSS for LCA includes \texttt{mclust} (Scrucca, Fop, Murphy, \& Raftery, 2016), \texttt{depmixS4} (Visser \& Speekenbrink, 2010), and \texttt{OpenMx} (Neale et al., 2016).
A crucial limitation of these existing packages is that some have limited functionality (e.g., \texttt{mclust}),
or implement specific innovations (e.g., \texttt{depmixS4}).
\texttt{OpenMx} is unique in that, like proprietary \emph{Mplus}, it implements LCA in the context of a fully featured structural equation modeling framework.
A major downside of \texttt{OpenMx} is that it is not very user-friendly and its functionality is poorly documented.

These limitations are addressed by the \texttt{tidySEM} package,
which provides a user-friendly interface for LCA in \texttt{OpenMx}.
The package also includes convenience functions to obtain relevant statistics, tests and plots in accordance with best practices in estimation and reporting of LCA, as described in this paper.
The primary purpose of the \texttt{tidySEM} package is to provide a tidy workflow for generating, estimating, reporting, and plotting structural equation models in a software-agnostic way.
The user interface is simple, as models can be specified using the widely used \texttt{lavaan} syntax (Rosseel, 2012).
The package contains user-friendly wrapper functions to conduct LCA with sensible defaults,
but the resulting models can still be fully customized making it a low threshold - high ceiling package.
For instance, LCA models constructed in \texttt{tidySEM} inherit the \texttt{MxModel} format from \texttt{OpenMx} which renders them compatible with various legacy functions from \texttt{OpenMx} and related LCA ecosystem .
Since \texttt{tidySEM} was developed with open science in mind, it facilitates fully-reproducible analyses and adheres to the \emph{FAIR (Findable, Accessible, Interoperable, and Reusable) software principles} (Lamprecht et al., 2020),
and the \href{https://bestpractices.coreinfrastructure.org/en/criteria}{\emph{OpenSSF Best Practices}} for FOSS software.

@@

LCA refers to methods for estimating unobserved
groups based on a parametric model of observed indicators of group
membership (Vermunt et al., 2004).
It has become popular across scientific fields,
under a number of different names;
most notably (\emph{finite Gaussian}) \emph{mixture modeling} and \emph{latent profile analysis}.
The concept of LCA can be understood in different ways.
Generally speaking, LCA assumes that the study population is composed of \(k\)
subpopulations.
It further assumes that the observed data are
a mixture of observations originating from those subpopulations;
hence the name \emph{mixture model}.
Consider the simplest possible
univariate model, which is a normal distribution.
This model has two parameters: the mean and the variance.
LCA aims to estimate the values of those parameters across \(k\) classes,
as well as estimate the probability of every individual observation \(i\) belonging to each of the \(1 \ldots k\) classes.
Commonly, the same model is fitted to all \(k\) classes,
but each class is characterized by a unique set of parameters for that model (i.e., class-specific means and variances).

@@

The distribution is evidently bimodal, which bodes well for the intended
mixture model. In this case, the number of classes is known a-priori.
When estimating a two-class mixture model, the detective observes that
the model estimates the mean shoe size of the two groups as equal to
9.22 and
7.25, which is close to the true
means of the two groups, namely 9.04
and 6.93. When tabulating estimated
group membership against observed (known) group membership, it can be
seen that women are classified with a high degree of accuracy, but men
are not:

@@

Another way to conceptualize LCA is through an analogy to a measurement model in structural equation modeling (Molenaar \& Eye, 1994).
A mixture model is like confirmatory factor analysis,
except for the fact that the latent variable is categorical instead of continuous.
Another difference between the two techniques is that factor analysis can group observed \emph{variables} into latent constructs, while LCA can group \emph{observations} into classes.
While factor analysis seeks to describe the relations between variables,
LCA seeks to describe unobserved groups and classify individual observations to those groups.
In line with this distinction, LCA is referred to as a ``person-centered'' technique, and
factor analysis as a ``variable-centered'' technique
(Masyn, 2013; Nylund-Gibson \& Choi, 2018).

When the focus is on the model parameters in each group, LCA can be
thought of as similar to a multi-group structural equation model. The
main distinction is that group membership is not known a-priori, but estimated from the data.
Whereas in a multi-group model, the data are split by group and treated
as independent samples, in LCA, all cases contribute to the estimation
of all parameters in all groups. The relative contribution of each case
to the parameters of each group is determined by that case's \emph{posterior
probability} of belonging to that group.

When the focus is on each individual's estimated class membership,
LCA can be thought of as a type of clustering algorithm that corrects for random measurement error by default, and is able to correct for systematic measurement error if it is included in the model.
Specifically, the posterior class probability that an individual belongs to a latent class
can be computed from the likelihood of that individual's observed data given the class-specific model.
These posterior class probabilities can be used to weight follow-up analyses,
or to assign individuals to classes based on their highest class probability.
If the classes are clearly distinct, measurement error is low, and the latter approach may be acceptable.
This perspective on mixture modeling is sometimes referred to as ``model-based
clustering'' (Hennig, Meila, Murtagh, \& Rocci, 2015; Scrucca et al., 2016).
Many clustering algorithms apply some recursive splitting algorithm to the data. By
contrast ``model-based'' clustering refers to the fact that LCA estimates
cluster membership based on a parametric model.
Since a parametric model estimates only a fixed number of parameters,
it can be a relatively parsimonious solution compared to non-parametric techniques.

@@

The term LCA was initially conceived to refer to analyses with categorical,
usually binary indicators (Collins \& Lanza, 2009).
It was later generalized to refer to any technique that estimates a categorical latent variable based on a
parametric model of observed indicators, as it only refers
to the purpose of the analysis, and does not imply restrictions to the
model used, or the level of measurement of the indicators (Vermunt et al., 2004).
To prevent ambiguity, we refer to LCA with binary or ordinal indicators as \emph{LCA with ordinal indicators},
and the sole \emph{LCA} expression as an umbrella term.
Given the abundance of redundant terms in use for closely related types of LCA models,
we here provide a rudimentary taxonomy.

A common type of LCA is the \emph{finite Gaussian mixture model}; a
univariate analysis where the observed distribution of a single variable
is assumed to result from a mixture of a known number of normal
(Gaussian) distributions. The parameters of a finite Gaussian mixture
model are the means and variances of these underlying normal
distributions. The analysis of shoe sizes presented earlier is a
canonical example of this type of analysis. In the multivariate case,
with more than one indicator variable, the parameters of a mixture model
are the means, variances, and covariances between the indicators (which
can be standardized to obtain correlations). These parameters can be estimated
freely, or set to be constrained across classes (Collins \& Lanza, 2009).

The technique known as \emph{latent profile analysis (LPA)} is a special case
of the mixture model, which assumes conditional independence of the
indicators. Conditional independence means that, after class membership
is accounted for,
the residual covariances/correlations between indicators are
assumed to be zero.
In some cases, such constraints
will be inappropriate; for instance when the cohesion between indicators
is expected to differ between classes (Collins \& Lanza, 2009).
Consider, for example, a mixture model of ocean plastic particles, which found two classes of particles based on length and width:
a class of larger particles with a low correlation between length and width,
and a class of smaller particles with a high correlation between length and width.
The reason for this difference in correlations makes theoretical sense:
the large particles were heterogenous in shape,
whereas the smaller particles had been polished to a more uniform (rounded) shape by the elements (Alkema, Van Lissa, Kooi, \& Koelmans, 2022).

A longitudinal extension of LCA is the \emph{Hidden Markov Model} (\emph{HMM}, also
called \emph{Latent Transition Analysis}), which estimates the change in class
membership over time under the Markov assumption. The Markov
assumption states that class membership at time \(t\) is only affected
by the class membership at time \(t-1\). The structural part of the model
is comprised of the initial states (proportion of the sample in each
class at the first time point), and transition probabilities
(probabilities of class membership switching across time). The transition
probabilities are estimated as multinomial logistic regressions between
class memberships (Vermunt, 2010).

Finally, some LCA models use latent variables as indicators.
A common application of this approach is in
longitudinal research with repeated measures of a construct.
Examples of this approach include
\emph{growth mixture models} (GMM) and \emph{latent class growth analyses} (LCGA)
(Jung \& Wickrama, 2008).
These techniques estimate a latent growth model
to describe grouped trajectories over time.
The indicators of class membership in a GMM are the
intercepts, variances, and covariances of the latent growth variables that describe individual trajectories on the observed variables.
For linear trajectories, these latent variables are the intercept and slope.
The free variances allow trajectories to vary within classes. The LCGA differs from the GMM in that it assumes within-class homogeneity of trajectories.
This is achieved by restricting the (co)variances of the growth parameters to zero.
Any variance in the indicators not explained by the class-specific
latent trajectories is assumed to be error variance.

@@

Although LCA is often discussed as an exploratory analysis technique,
it can also be used in a confirmatory manner.
Given that LCA is similar to confirmatory factor analysis,
it can be used to similar ends:
to assess whether a theoretical measurement model holds.
This use case is relevant when testing a theory that postulates the existence of a categorical latent variable.
For example, \emph{identity status theory} posits that, at any given point in time, adolescents reside in one of four identity statuses (Marcia, 1966).
LCA can be used to identify these four statuses based on observed indicators (e.g., self-reported identity exploration and commitment, Luyckx, Schwartz, Goossens, Soenens, \& Beyers, 2008).
Similarly, \emph{personality type theory} states that dimensional differences in personality can largely be explained by an undercontrolled, overcontrolled, and resilient personality types, which could also be restored using LCA (Donnellan \& Robins, 2010).

@@

Appendix A showcases a step-by-step tutorial to confirmatory LCA in \texttt{tidySEM},
based on a recent study that tested a theory that identities of Belgian adolescents with migration backgrounds could be summarized as those relating to their heritage, national and regional identities (Maene, D'hondt, Van Lissa, Thijs, \& Stevens, 2022).

@@

Another use case of LCA is to explore whether a population is comprised of latent subpopulations, and if so, how many.
The primary focus in this use case is on class enumeration.
For example,
one study explored heterogeneity in the population of military service members and veterans with prior suicide attempts,
and identified two latent profiles characterized by high versus low suicide risk based on self-reported suicidal ideation and prior attempt methods and severity
(Gromatsky et al., 2022).
Knowing that the population of suicidal service members is heterogeneous can inform future research, policy, and care.
Appendix B offers a tutorial for exploratory LCA,
based on the aforementioned study of the distribution of ocean microplastics (Alkema et al., 2022).

@@

LCA can also be used as a measurement model, similar to confirmatory factor analysis.
This is useful when class membership is measured imperfectly by several indicators:
LCA can partial out measurement error and restore most likely class membership.
For example, one study found discrepancies between self-reported employment status and official register data (Pavlopoulos \& Vermunt, 2015).
While it might be tempting to assume that register data is free of error, this assumption was shown to be false.
A mixture model was used to estimate employment status, correcting for measurement error, from both self-report and registry data.
The study investigated random and systematic measurement error in the indicatorss stemming from the two registers,
which revealed strengths and weaknesses of both types of assessment.

This example also illustrates that LCA allows us to evaluate the reliability of different types of indicators (Geiser \& Wurpts, 2014).
As a rule, a continuous indicator that discriminates well between classes will show large differences in class-specific means. An ordinal indicator with the same property will show very high or very low conditional response probabilities for some classes but not for others.
This information can be used to select the indicators that are most diagnostic of class membership.
For an example of LCA with a focus on differential item functioning,
see a study on university admission tests of Saudi Arabian students (Tsaousis, Sideridis, \& AlGhamdi, 2020).

@@

LCA can also be used to reduce high-dimensional data
to just a few prototypes.
Factor analysis is also sometimes used for dimension reduction,
but only accounts for linear covariance between indicators,
whereas LCA can accommodate complex - but discrete - patterns.
For example, imagine that there exist two indicators - \(x\) and \(y\). Some individuals score high on both indicators, some score low on both, and other score high on \(x\) but low on \(y\).
Factor analysis would only be able to capture the differences between the first two groups well, but LCA will capture differences between all three classes. Note, however, that this is a pragmatic use of LCA. There is no theory about the existence of a categorical latent variable (cf.~Testing theory).
Consequently, the final class solution should be treated as descriptive of the sample.
It should not be reified or treated as evidence of the existence of a categorical latent variable without accumulation of additional empirical evidence and deliberation on the findings. The discovered classes could indeed prove to be theoretically justified and thus inform the theory building process, but a single exploratory fining is never sufficient for reaching such conclusions.

The use of LCA for dimension reduction is common in longitudinal research,
where one developmental process is summarized using latent class growth analysis (LCGA),
and the resulting categories are then used as a moderator of another developmental process in a multi-group model.
For example, one study conducted a LCGA of adolescent empathy development on two dimensions of empathy identifying: high, average, and low empathy classes (Van Lissa et al., 2015).
The membership to these three groups was then used as a moderator in a latent growth analysis of adolescent- and parent-reported conflict.
The results showed that the high-empathy group evidenced greater adolescent-parent agreement about the incidence of conflict than other groups,
and the low empathy group had more conflict with parents than other groups.
Note that as both of these effects were non-linear,
modeling them in a single-group framework would introduce unnecessary bias.
LCGA greatly simplifies model specification and interpretation,
albeit at a cost of some loss of information, and potentially generalizability.
Appendix C is a tutorial on LCGA,
based on ongoing research on trajectories of depressive symptoms in military service members after deployment.

The need for dimension reduction can be especially pertinent when using ordinal indicators.
For example, one study measured
fifteen dichotomous health indicators among injured military personnel (MacGregor et al., 2021).
Combinatorics informs us that fifteen
dichotomous items have \(2^{15}\) or \(32,768\) unique symptom combinations.
However, LCA was able to reduce response patterns to five clinically meaningful latent classes supported by the data.

@@

Another use case of LCA is to estimate individuals' unobserved class membership based on observed indicators.
The shoe size example in Figure \ref{fig:shoedens}
is a rudimentary illustration of such application.
In applied research, LCA is often used for classification in clinical contexts.
For example, LCA was used to determine clinical cut-off scores for diagnosing whooping cough (Baughman, Bisgard, Lynn, \& Meade, 2006).
Another study used LCA not only to classify the sample used to estimate the model, but also predict future cases {[}REF Zegwaard et al, in prep{]}.
The study identified four types of care providers among those who supported close kin with mental health problems.
Importantly, the LCA model was embedded in an interactive app that could be used by healthcare professionals to determine the most likely class membership for new care providers and pave the way to providing them more targeted support.
To obtain predicted class membership based on an existing LCA model,
the authors computed the likelihood of data from a new participant under the multivariate normal distributions estimated within all latent classes.
These applications of LCA aid the clinical decision making process,
and help decide whether additional support is warranted,
or what type of intervention is appropriate.

@@

The matrix \(P\) can be summarized in different way.
For example, the estimated frequency table of the latent categorical variable can be obtained by summing the posterior probabilities by class (i.e., the column sums of \(P\)).
This frequency table \(F_P\) is obtained by calling \texttt{class\_prob(res,\ type\ =\ "sum.posterior"}.
These numbers are fractional because this table takes classification error into account.
Individuals can contribute partially to multiple classes.
